Verified Nanopasses for Compiling Conditionals

### 2025-08-12 paper.tex
fixes from Tianyu

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}
\noindent This idea builds on earlier work by \citet{Lawall:1994aa} on
continuation-based partial evaluation. Indeed, the above is a partial
evaluation of Boolean expressions.

@@

The third language replaces the \lstinline{Let} bindings with
assignment expressions but retains the structured control-flow and
expression-oriented nature of the language. This language brings us
closer to x86 where variables are replaced by registers and stack
locations, all of which are mutable. The abstract syntax for
assignment is \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the
right-hand side and $e_2$ is an expression to be executed after the
assignment (reminiscent of the body of a \lstinline{Let}).  The
\lstinline{Program} AST node consists of the number of variables and
an expression.

@@

Regarding the semantics, the main change is that environments are now
both an input and output. Also, in anticipation of the next language,
we switch to representing the semantics as a relation, that is, a
big-step semantics.

\begin{lstlisting}
StateImp = Inputs × Env Value
data _$⊢$_$⇓$_$⊣$_ : StateImp $→$ Imp-Exp
                     $→$ Value $→$ StateImp $→$ Set
\end{lstlisting}

\noindent Here is the rule for assignment:
\[
\inferrule{s , ρ ⊢ e_1 ⇓ v_1 ⊣ s′ , ρ′ \\
           s′ , ρ′(x := v_1) ⊣ e_2 ⇓ s″ , ρ″}
          {s , ρ ⊢ \mathtt{Assign}~ x~ e_1~ e_2 ⊣ s″ , ρ″}
\]

The semantics of whole programs is given by the following relation.

\begin{lstlisting}
interp-imp : Imp-Prog $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

@@

The base cases for \lstinline{explicate-pred} are interesting.  If the
expression is a literal \lstinline{true} or \lstinline{false}, then we
return the continuation for the \lstinline{thn} or \lstinline{els}
branch, respectively. (This may case \lstinline{thn} or
\lstinline{els} to become dead code, which we discuss later.)

>>>>

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code $c_1$
and $c_2$ for the two branches, that is, for the two continuations of
$b$.  Thus, $\mathcal{B}$ can analyze the expression $b$ to decide how
to generate code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is
even necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}
\noindent This idea builds on earlier work by \citet{Lawall:1994aa} on
continuation-based partial evaluation. Indeed, the above is a partial
evaluation of Boolean expressions.

@@

The third language replaces \lstinline{Let} bindings with assignment
expressions but retains the structured control-flow and
expression-oriented nature of the language. This language brings us
closer to x86 where variables are replaced by registers and stack
locations, which are mutable. The abstract syntax for assignment is
\lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the right-hand
side and $e_2$ is an expression to be executed after the assignment
(like the body of a \lstinline{Let}).  The \lstinline{Program} AST
node consists of the number of variables and an expression.

@@

Regarding the semantics, the main change is that environments are now
both an input and output. Also, in anticipation of the next language,
we switch to representing the semantics as a relation, that is, a
big-step semantics.
\begin{lstlisting}
StateImp = Inputs × Env Value
data _$⊢$_$⇓$_$⊣$_ : StateImp $→$ Imp-Exp
                     $→$ Value $→$ StateImp $→$ Set
\end{lstlisting}
\noindent Here is the rule for assignment:
\[
\inferrule{s , ρ ⊢ e_1 ⇓ v_1 ⊣ s′ , ρ′ \\
           s′ , ρ′(x := v_1) ⊣ e_2 ⇓ s″ , ρ″}
          {s , ρ ⊢ \mathtt{Assign}~ x~ e_1~ e_2 ⊣ s″ , ρ″}
\]
The semantics of programs is given by the following.
\begin{lstlisting}
interp-imp : Imp-Prog $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

@@

The base cases for \lstinline{explicate-pred} are interesting.  If the
expression is a literal \lstinline{true} or \lstinline{false}, then we
return the continuation for the \lstinline{thn} or \lstinline{els}
branch, respectively. (This may cause \lstinline{thn} or
\lstinline{els} to become dead code, which we discuss later.)
